---
title: "Final CA Analysis"
author: "Malcolm Chen"
date: "2025-12-02"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(scales)
library(broom)
library(knitr)

ca <- read_csv("datasets/clean_ca.csv")

ca <- ca %>%
  mutate(
    year = as.numeric(gsub("CY ", "", year)),
    utilization_rate = as.numeric(gsub("%", "", utilization_rate)) / 100,
    users = as.numeric(users),
    denominator = as.numeric(denominator),
    covid_period = ifelse(year >= 2020, 1, 0)
  )
```

# COVID-19 Interrupted Time Series (ITS)

This is the most important CA analysis.

We estimate the following model:

$$
\text{utilization}_{t}
= \beta_0
+ \beta_1 \cdot year_t
+ \beta_2 \cdot covid_t
+ \beta_3 \cdot (year_t \times covid_t)
$$

Where:

- **β₁**: pre-COVID slope  
- **β₂**: immediate level drop in 2020  
- **β₃**: slope change after COVID  

```{r covid_its_model}
its_model <- lm(
  utilization_rate ~ year + covid_period + year:covid_period,
  data = ca
)

tidy(its_model)
```

```{r covid_its_plot}
avg_year <- ca %>%
  group_by(year) %>%
  summarise(mean_util = weighted.mean(utilization_rate, denominator, na.rm = TRUE))

avg_year %>%
  ggplot(aes(year, mean_util)) +
  geom_line(linewidth = 1.2, color = "#2E86AB") +
  geom_point(size = 2) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "red") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "CA Dental Utilization – Interrupted Time Series (COVID-19)",
    subtitle = "Dashed line marks start of COVID in 2020",
    x = "Year", y = "Weighted Utilization Rate"
  ) +
  theme_minimal(base_size = 13)
```

### Interpretation
```{r}
its <- tidy(its_model)

cat("**Pre-COVID yearly slope:**", round(its$estimate[its$term=="year"],4), "\n\n")
cat("**Immediate drop at COVID (level change):**", round(its$estimate[its$term=="covid_period"],4), "\n\n")
cat("**Post-COVID slope change:**", round(its$estimate[its$term=="year:covid_period"],4), "\n\n")
```

The interrupted time-series analysis shows a clear and statistically significant upward trend in dental utilization prior to the COVID-19 pandemic (β₁ = 0.014, p < 0.001). In contrast, neither the estimated level change in 2020 (β₂ = 0.825, p = 0.961) nor the change in post-COVID slope (β₃ = –0.00043, p = 0.959) reached statistical significance. These results indicate that, although descriptive plots show an apparent decline during the onset of COVID-19, the regression model does not detect a statistically reliable interruption in either the level or trajectory of utilization. Overall, the data are most consistent with a continuation of the pre-existing upward trend, with no measurable long-term deviation attributable to the pandemic within this aggregated CA system-level dataset.











