---
title: "Clean_NHANES-Data"
author: Malcolm
output: 
  github_document
---

```{r}
library(tidyverse)
library(readr)
```

# Load the data
```{r}
nhanes <- read_csv("../datasets/nhanes_oral.csv")
```

# Delete useless cols
```{r}
cols_to_drop <- c("Presentation Standard", "Note1", "Note2", "Notea", "Noteb")
nhanes <- nhanes %>%
  select(-any_of(cols_to_drop))
```

# Rename cols
```{r}
nhanes <- nhanes %>%
  rename(
    survey_years = `Survey Years`,
    sex = Sex,
    age_group = `Age Group`,
    race_ethnicity = `Race and Hispanic Origin`,
    measure = Measure,
    percent = Percent,
    se = `Standard Error`,
    ci_lower = `Lower 95% CI Limit`,
    ci_upper = `Upper 95% CI Limit`
  )
```

# Clean percent
```{r}
nhanes <- nhanes %>%
  mutate(
    percent = str_replace_all(as.character(percent), "%", ""),
    percent = as.numeric(percent)
  )
```

# Clean years
#### When creating trend charts or performing regression analysis, a numerical time period is needed to represent the location of each point.
```{r}
nhanes <- nhanes %>%
  mutate(
    survey_years = str_replace_all(survey_years, "–", "-"),
    year_mid = as.numeric(str_sub(survey_years, 1, 4)) + 0.5
  )
```

# Standardize the Measure Name
```{r}
measure_map <- c(
  "Total Dental Caries in Primary Teeth" = "total_primary",
  "Total Dental Caries in Permanent Teeth" = "total_perm",
  "Untreated Dental Caries in Primary Teeth" = "untreated_primary",
  "Untreated Dental Caries in Permanent Teeth" = "untreated_perm",
  "Sealants on Permanent Teeth" = "sealant_perm",
  "Sealants on Permanent and Primary Teeth" = "sealant_all"
)

nhanes <- nhanes %>%
  mutate(measure_clean = recode(measure, !!!measure_map))
```

# No changes have been made to the groupings at present (except for retaining only age groups instead of larger groups).
```{r}
nhanes <- nhanes %>%
  filter(age_group %in% c(
    "2-5", "6-11", "12-19",
    "20-29", "30-39", "40-49", "50-59",
    "60-69", "70 and over"
  ))
```

# Delete the rows that missing percent
```{r}
nhanes_clean <- nhanes %>%
  filter(!is.na(percent))
```

# Sort and output
```{r}
nhanes_clean <- nhanes_clean %>%
  select(
    survey_years, year_mid, sex, age_group,
    race_ethnicity, measure, measure_clean,
    percent, se, ci_lower, ci_upper
  )
```

# Save
```{r}
write_csv(nhanes_clean, "../datasets/nhanes_oral_clean.csv")
```








